### 2024-07-23 Paper/paper.tex

    \begin{abstract}
		{\small
        There is mixed evidence on whether not-for-profit hospitals differ meaningfully from for-profit (FP) hospitals, but what has not been considered is the effect leadership teams have on hospital behavior. In this paper, I investigate whether leadership team composition of NFP hospitals affects how similarly they behave to for-profits. I use a novel data set on NFP hospital executives in the US (2010-2014) to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage pay-for-performance initiatives in the US enacted in 2012, and measure the  readmission and mortality rate responses between FP and NFP with and without clinically trained executives. I find that for-profits nad NFPs without clinical executives respond more aggressively to financial incentives on quality, consistent with theoretical predictions of firms that value profit relatively more than having high quality. 
		} 
	\end{abstract}

@@

@@

  Using publicly available tax forms, I construct a novel data set of NFP hospital executives from 2009-2015. I link hospitals in this data to characteristics in the American Hospital Association (AHA) survey and Hospital Compare data. I leverage various pay-for-performance initiatives in 2012 to investigate differential hospital responses between for-profits, and NFPs with and without without clinically trained executives.

@@

    The historical tax forms for all not-for-profits are housed by ProPublica.\footnote{https://projects.propublica.org/not-for-profits/} I use the NonProfit Explorer API to extract Employee Identification Numbers (EIN) of NFPs with a hospital National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are approximately 3,000 NFP EINs. Linking these EINs to other hospital information is crucial for the analysis. I therefore rely on matching by location and name of hospital in both the tax forms and AHA data. I confidently match 1200 EINs to an AHA identification number based on exact name matches within the same state. I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.

@@

    In Table \ref{tab:sumstats_samples}, I present a table of means of relevant variables for several hospital samples used throughout the analysis. First, there are 708 for-profit hospitals in the sample, shown in the left-most column. Notably,

    There are several samples of hospitals that are relevant for the analysis. First, there are all for-profits and all not-for-profits. Further, I subset the sample of NFPs into whether they have a clinical executive at some point in the sample. Finally, I characterize a NFP leadership team as being stable or not. A stable team means the hospital did not change their propensity to hire a clinical executive over time. I present a table of means for each of these sub-samples in Table \ref{tab:sumstats_samples}. Approximately 850 NFPs have leadership information, with 384 of them hiring an MD executive at some point in the sample. I compare these NFPs with the 708 for-profits from the AHA Survey. For-profits and NFPs are roughly the same bed size, but NFPs see more patients in each of the three relevant conditions. For-profits are more likely to be penalized, but also more likely to receive HVBP incentives (potentially for improvement in quality). For-profits have higher readmission rates, and higher mortality rates in all categories except heart failure. There are also key differences in not-for-profits with and without clinically trained executives. Those with MDs are larger, more likely to be penalized, have higher uncompensated care, and more complex patients. This holds for stable leadership teams as well, though the number of NFPs with a clinical executive drops significantly. I discuss the implications of these differences for identification in Section \ref{sec:model}.

@@

    I model hospital behavior under two simplified worlds: one where quality does not directly affect profit, and one where it does. Pay-for-performance incentives essentially incorporate performance directly into a hospital's profit function. I directly model performance as quality of care, which is inversely related to readmission and mortality rates. While the literature speculates about the true form of NFP objective functions, we can generally think of hospitals choosing some linear combination of profit and societal benefit, without knowing how much weight the hospital places on each. Intuitively, a hospital that places only weight on profit (for-profit type) will only care about quality when it enters into the profit function, and therefore will respond more drastically to pay-for-performance than a hospital who already cared about quality through societal benefit before the incentives. Therefore, hospital responses to pay-for-performance incentives reveal information about their underlying preference on profit vs. societal outcomes.

@@

    \noindent where $R$ is net revenue, $c_{\theta}(.)$ is an increasing cost function specific to quality $\theta$, $u(.)$ is extra utility gained from quality, which is increasing and concave in $\theta$, and $\alpha\in[0,1]$ captures the weight placed on profit vs. societal benefit. There is an implicit stay open condition throughout the model. Taking the first order condition yields

@@

    In light of the theoretical predictions, I test empirically whether different hospital types respond to pay-for-performance incentives differently. I present here a two-way fixed effects specification, where the key variable is an interaction between a certain hospital type and time periods after program enactments. Ultimately, I employ synthetic difference-in-differences, which is estimating the same model with weights that more heavily factor similar trending control groups and time periods. However, for clarity, I present the unweighted specification:

@@

    An important distinction is whether the classification of NFPs by leadership team is time-varying. For overall ownership status, switching is rare, and I limit to those who never switch. However, hospital executive teams are dynamic, and whether or not a hospital employs a clinically trained executive in each period may change. Therefore, I first present effects of having a clinically trained executive among stable executive teams, i.e., NFPs with no change in their propensity to hire a clinically trained executive.

>>>>

    \begin{abstract}
		{\small
        In highly specialized careers, it is not uncommon to have leaders with specific training, such as lawyers becoming judges, athletes becoming coaches, or doctors becoming hospital leaders. However, there is little evidence on how specialized training vs. general leadership training affects firm behavior. In this paper, I investigate empirically the effect of clinically trained executives on hospital behavior. In particular, I leverage policies changing quality incentives for hospitals in the US in 2012, and estimate the difference in readmission and mortality rate responses for nonprofit hospitals with and without clinically trained executives. I find that a lack of clinical executives leads hospitals to improve quality more drastically after the incentive change. Further, I provide evidence that the more drastic response is due to non-clinical executives being more profit driven, and that clinical executives are not just a signal of underlying hospital objectives, but manage the hospital differently. 
		} 
	\end{abstract}

@@

    There are many industries where some leaders have highly specialized training specific to that industry. For example, lawyers can be promoted to judges, athletes can be promoted to coaches, or doctors can be promoted to hospital leaders. However, there is little evidence on how specialized vs. general training in leadership affects firm behaviors. In this paper, I investigate how having clinically trained executives in leadership affects hospital behavior.

@@

  Using publicly available tax forms, I construct a novel data set of NFP hospital executives from 2009-2015, identifying which executives have clinical training. I link hospitals in this data to characteristics in the American Hospital Association (AHA) survey and Hospital Compare data. I leverage various pay-for-performance initiatives in 2012 to investigate differential hospital responses between for-profits, and NFPs with and without without clinically trained executives.

@@

    The historical tax forms for all not-for-profits are housed by ProPublica.\footnote{https://projects.propublica.org/not-for-profits/} I use the NonProfit Explorer API to extract Employee Identification Numbers (EIN) of NFPs with a hospital National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are approximately 3,000 NFP EINs. Linking these EINs to other hospital information is crucial for the analysis. I therefore rely on matching by location and name of hospital in both the tax forms and AHA data. In the AHA data, I limit to general hospitals. Then, I confidently match 1200 EINs to an AHA identification number based on exact name matches within the same state. I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.

@@

    In Table \ref{tab:sumstats_samples}, I present a table of means of relevant variables for several hospital samples used throughout the analysis. First, there are 708 for-profit hospitals in the sample, shown in column (1). For-profit hospitals in the sample have 155 beds on average, 112-276 patients depending on the condition, and are highly likely to be affiliated with a system. Thirty six percent of for-profits were penalized for all three conditions at some point.

@@

    I model hospital behavior under two simplified worlds: one where quality does not directly affect profit, and one where it does. Pay-for-performance incentives essentially incorporate performance directly into a hospital's profit function (\cite{dranove2011health}). I directly model performance as quality of care, which is inversely related to readmission and mortality rates. While the literature speculates about the true form of NFP objective functions, we can generally think of hospitals choosing some linear combination of profit and societal benefit, without knowing how much weight the hospital places on each. Intuitively, a hospital that places only weight on profit (for-profit type) will only care about quality when it enters into the profit function, and therefore will respond more drastically to pay-for-performance than a hospital who already cared about quality through societal benefit before the incentives. Therefore, hospital responses to pay-for-performance incentives reveal information about their underlying preference on profit vs. societal outcomes.

@@

    \noindent where $R$ is net revenue, $c_{\theta}(.)$ is an increasing cost function specific to quality $\theta$, $u(.)$ is extra utility gained from quality, which is increasing and concave in $\theta$, and $\alpha\in[0,1]$ captures the weight placed on profit vs. societal benefit. There is an implicit stay-open condition throughout the model. Taking the first order condition yields

@@

    In light of the theoretical predictions, I test empirically whether different hospital types respond to pay-for-performance incentives differently. I present here a two-way fixed effects specification, where the key variable is an interaction between a certain hospital type and time periods after program enactments. Ultimately, I employ synthetic difference-in-differences, which is effectively estimating the same model with weights that more heavily factor similar trending control groups and time periods. However, for clarity, I present the unweighted specification:

@@
